---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  dpi = 400,
  out.width = "100%",
  out.extra = 'style="border: 1px solid darkgray;"'
)
```

# grid.table.minimal

<!-- badges: start -->
<!-- badges: end -->

The goal of grid.table.minimal is to quickly format grid graphics based tables and plots.

## Installation

You can install the development version of grid.table.minimal from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("DannyMRoss/grid.table.minimal")
```

## Examples

```{r example}
library(grid.table.minimal)
library(dplyr)
library(ggplot2)
library(data.table)
library(extrafont)
library(showtext)
library(grid)
library(gridExtra)
library(gtable)
set_border_defaults()
loadfonts()
```

```{r include=F}
fonts <- c(paste0(getwd(),"/fonts/ttf"))
options("fontsize" = 8)
options("border.fontsize" = 10)
iris <- copy(iris)
```

#### summary table for examples
```{r tablesetup}
setDT(iris)
DT <- iris[sample(.N, 10), 1:4]
DT <- rbind(DT, 
            DT[, lapply(.SD, function(x) paste0(mean(x), " (",round(sd(x),2),")"))])
setnames(DT, rep(c("Length","Width"),2))
print(DT)
```

## Example 1
### grid.table.minimal() converts a data.frame to a gtable object with minimalist formatting applied. The gtable can then be drawn to the graphics device with grid.draw()
```{r example1}
g <- grid.table.minimal(DT)
grid.draw(g)
```

### grid.borders() draws borders, title, subtitle, caption, and watermark to the graphics device
```{r exampleborders}
grid.borders(
  bottom = c(l="BOTTOM LEFT", c="BOTTOM CENTER", r="BOTTOM RIGHT"), 
  top = c(l="TOP LEFT", c="TOP CENTER", r="TOP RIGHT"),
  left = c(b="LEFT BOTTOM", c="LEFT CENTER", t="LEFT TOP"),
  right = c(b="RIGHT BOTTOM", c="RIGHT CENTER", t="RIGHT TOP"),
  title = "Title",
  subtitle = "Subtitle",
  caption = list("Caption" = c("[1]")),
  watermark = "Watermark"
)
```

```{r include=FALSE}
invisible(dev.off()) # clear device
```

### draw borders over the gtable
```{r borderdraw}
grid.draw(g)
grid.borders(title = "Minimal Table Example")
```

```{r include=FALSE}
invisible(dev.off()) # clear device
```

### grid.table.minimal() and grid.borders() rely on global options that are set to default values by the below function.
```{r borderdefaults}
set_border_defaults(show = TRUE)
```

### example save to pdf usage
```{r savetable}
set_border_defaults(plot.width = 8.5, plot.height = 11, fontsize=16)
pdf("man/figures/table1.pdf", width = getOption("plot.width"), height = getOption("plot.height")) # start pdf stream
grid.slip(label = "Ex. 1") # draws a slip sheet
grid.draw(grid.table.minimal(DT))
grid.borders(title = "Minimal Table Example", bottom = c(l="",c="",r="Ex. 1"))
invisible(dev.off()) # close pdf stream
```

```{r include=FALSE}
embedFonts("man/figures/table1.pdf", outfile = "man/figures/table1.pdf", fontpaths = fonts) # in case user does not have fonts installed
```

[man/figures/table1.pdf](man/figures/table1.pdf)

```{r include=F}
set_border_defaults(border.fontsize=10, fontsize=8)
```

## Example 2
### gtables can be easily and flexibly formatted with the below functions
These functions extend those presented in [gridExtra/vignettes/tableGrob.html](https://cran.r-project.org/web/packages/gridExtra/vignettes/tableGrob.html) 
```{r example2}
# use grid.table.minimal() to convert DT to gtable object w/ minimal styling applied
g <- grid.table.minimal(DT)
# add column header spans with gtable_add_cspan()
g <- gtable_add_cspan(g, text="Sepal", r=2) # defaults to adding above 
g <- gtable_add_cspan(g, text="Petal", t=1, b=1, l=3, r=4) # specify t=b=1 to add to the existing top row
g <- gtable_add_cspan(g, text="Iris Random Sample (N = 10)")
# add horizontal lines with gtable_add_hline()
g <- gtable_add_hline(g, rows = nrow(g)-1) # adds hline(s) below row(s)
# format rows/cols/cells with grid.table.gp()
g <- grid.table.gp(g, rows = nrow(g), bold=T) # bold quickly with bold parameter
# individual cell formatting
g <- grid.table.gp(g, 
                   cols = c(1,3), 
                   rows = nrow(g), 
                   union = FALSE, # only apply to intersection of cols/rows
                   name = 'core-bg', # modify background grobs
                   gp_params = list(fill = "#DEEBF7"))
# draw gtable to device
grid.draw(g)
# draw borders, titles, caption, to device
grid.borders(title = "Minimal Table Example",
             subtitle = "flexible formatting",
             caption = list("Notes" = c("Summary row is the sample mean (sd).")),
             bottom = c(l="",c="",r="Ex. 2"), # add bottom margin text
             top = c(l = "", c="", r="")) # add top margin text
```

```{r include=FALSE}
invisible(dev.off()) # clear device
```

## Example 3
```{r include=F}
cl <- t(as.data.table(paste0("[",LETTERS[1:4],"]")))
colnames(cl) <- names(DT)
DT <- rbind(cl, DT)
```
### shortcuts are available to white-on-black the column headers
```{r example3}
set_border_defaults(default.fontfamily="Arial", fontsize=8, border.fontsize=10) # change default font and fontsize
g <- grid.table.minimal(DT, 
                        width_pct = .5, # pct of page that the table will take up
                        black.heading = TRUE) # black heading option
g <- gtable_add_cspan(g, text="Sepal", r=2, black = TRUE) 
g <- gtable_add_cspan(g, text="Petal", t=1, b=1, l=3, r=4, black = TRUE)
g <- gtable_add_hline(g, rows = nrow(g)-1)
g <- grid.table.gp(g, rows = nrow(g), bold=T)
grid.draw(g)
grid.borders(title = "Minimal Table",
             subtitle = "Black Headers",
             caption = list("Notes & Sources:" = c("Summary row is the sample mean (sd).")),
             bottom = c(l="",c="",r="Ex. 3"),
             top = c(l = "", c="", r=""))
```

```{r include=FALSE}
invisible(dev.off()) # clear device
```

# theme_borders()
### Use ggplot theme_borders() to add minimalist formatting to ggplots and set margins compatible with grid.borders()
Space above the plot is preserved for the title, subtitle, and caption labels to be added with `grid.borders()`
```{r pex1, fig.height=8.5, fig.width=11}
set_border_defaults()
p <- ggplot(iris, aes(x=Sepal.Width, y=Sepal.Length, color=Species)) + 
  geom_point() + labs(color="") +
  theme_borders()
grid.draw(p)
```

```{r include=FALSE}
invisible(dev.off()) # clear device
```

### Add titles, caption, and borders
```{r pex1tcb, fig.dim = c(11,8.5)}
grid.draw(p)
grid.borders(
  titlebox.draw = TRUE,
  title = "Title",
  subtitle = "Subtitle",
  caption = list("Caption" = c("[1]", "[2]")),
  bottom = c(l="", c="", r="Ex. 1")
)
```


